Solving the maximum independant set problem with einsum networks

### 2021-06-02 paper/paper.tex
get single configuration

\author{Jin-Guo Liu\\
Harvard University\\
\texttt{jinguoliu@g.harvard.edu}\\
\AND
Xun Gao\\
Harvard University\\
\texttt{xungao@g.harvard.edu}\\
\AND
Sheng-Tao Wang\\
QuEra computing Inc.
}
\maketitle

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is much easier. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= -\infty\\
    \mymathbb{1} &= 0
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$. This is exactly the tropical tensor network~\cite{Liu2021} used in solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$, we can only consider the powers, which corresponds to the tropical algebra~\cite{Maclagan2015}.

One may want to obtain all ground state configurations,
it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$ initialized to a vertex index $i$ dependent onehot vector $s_i = \{\boldsymbol{\delta}_{ij=1,\ldots,n}\}$.
The bit strings has a redefined algebra
\begin{align}
\begin{split}
    s_x \oplus s_y &= s_x \cup s_y\\
    s_x \odot s_y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{\boldsymbol 0\}
\end{split}
\end{align}
The $\vee$ is the bit-wise \texttt{or}. One can easily check that this replacement does not change the fact that the algebra is a commutative semiring.

@@

\end{document}

>>>>

\author{Jin-Guo Liu\\
Harvard University\\
\texttt{jinguoliu@g.harvard.edu}\\
\AND
Xun Gao\\
Harvard University\\
\texttt{xungao@g.harvard.edu}\\
}
\maketitle

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is much easier. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$. This is exactly the tropical tensor network~\cite{Liu2021} used in solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$, we can only consider the powers, which corresponds to the tropical algebra~\cite{Maclagan2015}.

One may want to obtain all ground state configurations,
it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$ initialized to a vertex index $i$ dependent onehot vector $s_i = \{\boldsymbol{\delta}_{ij=1,\ldots,n}\}$.
\iffalse
\begin{align}
\begin{split}
    s_x \oplus s_y &= s_x \cup s_y\\
    s_x \odot s_y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{\boldsymbol 0\}
\end{split}
\end{align}
\fi
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y}\\
    \mymathbb{0} &= \{\}\infty^{-\infty}\\
    \mymathbb{1} &= \{\boldsymbol 0\}\infty^{0}
\end{split}
\end{align}
The $\vee$ is the bit-wise \texttt{or}. One can easily check that this replacement does not change the fact that the algebra is a commutative semiring.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save computational effort.
We leave this as an exercise for readers.

@@

\section{Indenpendence polynomials}
\subsection{Common}
\begin{lstlisting}
using Polynomials

_auto_mispolytensor(x::T, ix::NTuple{2}) where T = T[1 1; 1 0]
_auto_mispolytensor(x::T, ix::NTuple{1}) where T = T[1, x]
function generate_polyxs!(x::T, code::OMEinsum.NestedEinsum, xs=Vector{Any}(undef, ninput(code))) where {T}
    for (ix, arg) in zip(OMEinsum.getixs(code.eins), code.args)
		if arg isa Integer
			xs[arg] = _auto_mispolytensor(x, ix)
		else
        	generate_polyxs!(x, arg, xs)
		end
    end
	return xs
end

function generate_polyxs!(x::T, code::EinCode, xs=Vector{Any}(undef, ninput(code))) where {T}
    for (i,ix) in enumerate(OMEinsum.getixs(code))
		xs[i] = _auto_mispolytensor(x, ix)
    end
	return xs
end

function mis_polysolve(code, x::T) where {T}
	xs = generate_polyxs!(x, code, Vector{Any}(undef, NoteOnTropicalMIS.ninput(code)))
	code(xs...)
end
\end{lstlisting}

\subsection{Symbolic}
\begin{lstlisting}
function independence_polynomial(::Val{:polynomial}, code)
    mis_polysolve(code, Polynomial([0, 1.0]))[]
end
\end{lstlisting}

\subsection{Fitting and FFT}
\begin{lstlisting}
using FFTW
function independence_polynomial(::Val{:fft}, code; mis_size=Int(mis_solve(code)[].n), r=1.0)
	ω = exp(-2im*π/(mis_size+1))
	xs = r .* collect(ω .^ (0:mis_size))
	ys = [mis_polysolve(code, x)[] for x in xs]
	Polynomial(real.(ifft(ys) ./ (r .^ (0:mis_size))))
end

function independence_polynomial(::Val{:fitting}, code; mis_size=Int(mis_solve(code)[].n))
	xs = (0:mis_size)
	ys = [mis_polysolve(code, x)[] for x in xs]
	fit(xs, ys, mis_size)
end
\end{lstlisting}

\subsection{Finite field algebra}
\begin{lstlisting}
using Mods, Primes

# pirate
Base.abs(x::Mod) = x
Base.isless(x::Mod{N}, y::Mod{N}) where N = mod(x.val, N) < mod(y.val, N)

function _independance_polynomial(::Type{T}, code, mis_size::Int) where T
	xs = 0:mis_size
	ys = [mis_polysolve(code, T(x))[] for x in xs]
	A = zeros(T, mis_size+1, mis_size+1)
	for j=1:mis_size+1, i=1:mis_size+1
		A[j,i] = T(xs[j])^(i-1)
	end
	A \ T.(ys)
end

function independence_polynomial(::Val{:finitefield}, code; mis_size=Int(mis_solve(code)[].n), max_order=100)
    N = typemax(Int)
    YS = []
    local res
    for k = 1:max_order
	    N = prevprime(N-1)
        T = Mods.Mod{N,Int}
        rk = _independance_polynomial(T, code, mis_size)
        push!(YS, rk)
        if max_order==1
            return Polynomial(Mods.value.(YS[1]))
        elseif k != 1
            ra = improved_counting(YS[1:end-1])
            res = improved_counting(YS)
            ra == res && return Polynomial(res)
        end
    end
    @warn "result is potentially inconsistent."
    return Polynomial(res)
end

function improved_counting(sequences)
    map(yi->Mods.CRT(yi...), zip(sequences...))
end
\end{lstlisting}

\section{Enumerating configurations}
\begin{lstlisting}
using OMEinsum, TropicalNumbers

struct ConfigEnumerator{N}
    data::Vector{BitVector}
end

Base.length(x::ConfigEnumerator{N}) where N = length(x.data)
Base.:(==)(x::ConfigEnumerator, y::ConfigEnumerator) = x.data == y.data

function Base.:+(x::ConfigEnumerator{N}, y::ConfigEnumerator{N}) where N
    res = ConfigEnumerator{N}(vcat(x.data, y.data))
    @show length(x), length(y), length(res)
    @assert length(res) == length(x) + length(y)
    return res
end

function Base.:*(x::ConfigEnumerator{L}, y::ConfigEnumerator{L}) where L
    M, N = length(x), length(y)
    z = Vector{BitVector}(undef, M*N)
    for j=1:N, i=1:M
        z[(j-1)*M+i] = x.data[i] .| y.data[j]
    end
    #@show M, N, length(z)
    @assert length(z) == length(x) * length(y)
    return ConfigEnumerator{L}(z)
end

Base.zero(::Type{ConfigEnumerator{N}}) where N = ConfigEnumerator{N}(BitVector[])
Base.one(::Type{ConfigEnumerator{N}}) where N = ConfigEnumerator{N}([falses(N)])

function onehot(::Type{ConfigEnumerator{N}}, i::Int) where N
    res = one(ConfigEnumerator{N})
    res.data[1][i] |= true
    return res
end

function enumerator_t(::Type{T}, ix::NTuple{2}, vertex_index) where {T1, N, T<:CountingTropical{T1, ConfigEnumerator{N}}}
    [one(T) one(T); one(T) zero(T)]
end
function enumerator_t(::Type{T}, ix::NTuple{1}, vertex_index) where {T1, N, T<:CountingTropical{T1, ConfigEnumerator{N}}}
    [one(T), CountingTropical(one(T1), onehot(ConfigEnumerator{N}, vertex_index[ix[1]]))]
end

symbols(::EinCode{ixs}) where ixs = unique(Iterators.flatten(ixs))
symbols(ne::NestedEinsum) = symbols(Iterators.flatten(ne))
function mis_enumerate(code)
    vertex_index = Dict([s=>i for (i, s) in enumerate(symbols(code))])
	xs = generate_xs!((T, ix)->enumerator_t(T, ix, vertex_index), CountingTropical{Float64,ConfigEnumerator{length(vertex_index)}}, code, Vector{Any}(undef, ninput(code)))
	code(xs...)
end
\end{lstlisting}
\end{document}
